---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PButils

## Overview

**PButils** provides utility functions to support the processing of planetary boundary data extracted from the literature.

A common issue in planetary boundary and footprint datasets is that **values and units are stored together in the same column**, making comparison, harmonisation, and calculation difficult.

The first core functionality of PButils is to **separate numeric values from their units**, while preserving the original data.

Future functionality will focus on:

* unit harmonisation across comparable variables,
* calculation of boundary overshoot at territorial and per-capita scales,
* visualisation helpers (e.g. cartograms).

## Installation

You can install the development version of PButils from [GitHub](https://github.com/) with:

```{r intall, eval = FALSE}
# install.packages("devtools")
devtools::install_github("GiorgiaGraells/PButils")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r loadPBUTILS}
library(PButils)
## basic example code
```

## Example data

PButils ships with an example dataset used throughout the package documentation:

```{r Nitrogendataexample}
data("Nitrogen_Data")

names(Nitrogen_Data)
```

This dataset contains several columns where numeric values and units are stored together as character strings (e.g. `"5.4 kg N cap⁻¹ yr⁻¹"`).

## Separating values and units

### Single column

The `separate_unit_value()` function separates a column containing both values and units into two new columns:

* `<column>_value` (numeric)
* `<column>_unit` (character)

while **preserving the original column**.

```{r showseparateunitvaluefunc}
out <- separate_unit_value(
  df = Nitrogen_Data,
  columns = "boundary_percapita"
)

names(out)
```

```{r whatIsanOut}
head(
  out[, c("boundary_percapita",
          "boundary_percapita_value",
          "boundary_percapita_unit")]
)
```

### Multiple columns at once

You can also process multiple columns in a single call:

```{r examplesofseparate}
out2 <- separate_unit_value(
  df = Nitrogen_Data,
  columns = c("global_limit_considered", "boundary_percapita")
)

names(out2)
```

Each input column generates its own `_value` and `_unit` pair:

```{r showdata}
head(
  out2[, c("global_limit_considered_value",
           "global_limit_considered_unit",
           "boundary_percapita_value",
           "boundary_percapita_unit")]
)
```

## Low-level helpers

PButils also exposes the underlying helper functions used internally:

```{r examplesofExtract}
extract_value(Nitrogen_Data$boundary_percapita)
extract_unit(Nitrogen_Data$boundary_percapita)
```

These can be useful when working directly with vectors rather than data frames.

## Cartogram visualisation

PButils includes a helper to visualise planetary-boundary overshoot as a cartogram (country areas resized by overshoot or a transformation of it).

### Example

This example uses the bundled datasets `PBData` (overshoot values) and `World` (country polygons).

```{r ClimateChangePlot}
# Load example data
# (If you attached these datasets to the package, they will be available via data())
data("PBData")
data("World")

# Create a cartogram for a chosen boundary
res <- pb_cartogram(
  PB_df = PBData,
  world_sf = World,
  boundary = "Climate change",
  count_mode = "overshoot"
)

# Print the plot
res$plot
```

### What you get back

`pb_cartogram()` returns a list with:

* `carto_sf`: cartogram geometries (sf)
* `top_labels_sf`: centroid points for the labeled countries (sf)
* `plot`: a ggplot cartogram
* `data_joined`: input polygons joined to PB data (sf)
* `data_ea`: equal-area projected data used for area calculations (sf)

## Notes

* Original columns are **never modified or removed**
* Newly created _value and _unit columns are inserted immediately after their source column to facilitate quick inspection.
* Missing values are preserved as `NA`
* The implementation avoids non-base piping to keep dependencies minimal

## Development status

PButils is under active development. The API may evolve as additional unit harmonisation and overshoot calculations are added.

Bug reports and feature requests are welcome: [https://github.com/GiorgiaGraells/PButils/issues](https://github.com/GiorgiaGraells/PButils/issues)
